### 2016-10-04 Chapters/SolvationAndCorrelation.tex
DONE

% mainfile: ../RobertoDiRemigioPhDThesis.tex
%************************************************
\chapter{Electron Correlation and Solvation}\label{ch:solvation-correlation}
\footnotetext{Potius sero quam numquam \\
\hspace*{15pt} --- \textsc{Titus Livius}, \\ \hspace*{15pt}
\textit{Ab Urbe Condita}}

@@

The first approximate scheme truncates the source term to its

@@

   \denom{\mu_q}\tamp{\mu_q} + \braket{\mu_q | \overline{\Phi} | \mathrm{HF}}
   + {}^t\p\braket{\mu_q | \check{\s} | \mathrm{HF}}\delta_{\mu_q\mu_1}

@@

   \denom{\mu_q}\tamp{\mu_q} + \braket{\mu_q | \overline{\Phi} | \mathrm{HF}}
   + {}^t\p\braket{\mu_q | \check{\s} | \mathrm{HF}}\delta_{\mu_q\mu_1}

@@

Both the amplitudes and multipliers equations are formally identical to

@@

The \acrshort{PTES} approximation can be brought one step further, by
truncating the \acrshort{CCS} expectation value to its multiplier-independent
term already in the effective Lagrangian.
The coupling between the amplitudes and multipliers is broken, while
still improving upon the reference reaction field with correlated
contributions.\autocite{Caricato2011-tx}
The effective Lagrangian would then read as:
\begin{equation}\label{eq:pte-s-cc}
    {}^{\mathrm{PTE(S)}}\lag(\tamp{}, \tbar{}, \p)_\mathcal{M} =
  {}^\mathrm{PTE}\lag(\tamp{}, \tbar{})_\mathcal{M}
  + \frac{1}{2}{}^t\p\V\p + {}^t\p\braket{\mathrm{HF} | \check{\s}_{\mathrm{N}} | \mathrm{HF}}
\end{equation}
which uncouples all equations:
\begin{subequations}
  \begin{align}
  {}^{\mathrm{PTE(S)}}\tampEq{\mu_q}(\tamp{}, \tbar{})  &=
   \denom{\mu_q}\tamp{\mu_q} + \braket{\mu_q | \overline{\Phi} | \mathrm{HF}}
   = 0 \\
   {}^{\mathrm{PTE(S)}}\tbarEq{\mu_q}(\tamp{}, \tbar{}, \p) &=
    \denom{\mu_q}\tbar{\mu_q} +
    \braket{\mathrm{HF} | \commutator{\overline{\Phi}}{\cluster{\mu_q}} | \mathrm{HF}} +
    \sum_{u=1}^{\mathcal{M}}\braket{\tbar{u} |
    \commutator{\overline{\Phi}}{\cluster{\mu_q}} | \mathrm{HF}}
    \nonumber \\
    &+
    {}^t\p\braket{\mathrm{HF} | \commutator{\s}{\cluster{\mu_q}} | \mathrm{HF}}\delta_{\mu_q\mu_1}
    = 0 \\
    {}^{\mathrm{PTE(S)}}\Omega_\p(\tamp{}, \tbar{}, \p)
    &=
    \V\p + \braket{\mathrm{HF} | \check{\s}_{\mathrm{N}} | \mathrm{HF}} = 0
    \label{eq:pte-s-polarization}
  \end{align}
\end{subequations}
The amplitude equations are formally equivalent to those derived in the
\acrshort{PTE} scheme, \emph{vide infra}.
The equations for the multipliers are also formally equivalent to the
\acrshort{PTE} equations, apart from the equation determining the
singles multipliers which features an additional term involving the
polarization:
${}^t\p\braket{\mathrm{HF} | \commutator{\s}{\cluster{\mu_q}} | \mathrm{HF}}\delta_{\mu_q\mu_1}$.
This term is not, however, coupled to the other equations given that
$\p$ is fixed once the singles amplitudes have been determined.

@@

Complete neglect of the polarization functional leads to the most
drastic approximation to the full quantum/classical polarizable
effective Lagrangian, the \acrshort{PTE} scheme.
This is equivalent to differentiating the polarization-independent
\acrshort{PTE} effective Lagrangian:
\begin{subequations}
  \begin{align}
 {}^{\mathrm{PTE}}\tampEq{\mu_q}(\tamp{}, \tbar{}, \p)  &=
   \denom{\mu_q}\tamp{\mu_q} + \braket{\mu_q | \overline{\Phi} | \mathrm{HF}}
    = 0 \\
   {}^{\mathrm{PTE}}\tbarEq{\mu_q}(\tamp{}, \tbar{}, \p)
    &=
    \denom{\mu_q}\tbar{\mu_q} +
    \braket{\mathrm{HF} | \commutator{\overline{\Phi}}{\cluster{\mu_q}} | \mathrm{HF}} +
    \sum_{u=1}^{\mathcal{M}}\braket{\tbar{u} |
    \commutator{\overline{\Phi}}{\cluster{\mu_q}} | \mathrm{HF}}
    = 0
  \end{align}
\end{subequations}
The \acrshort{PTE} scheme naturally preserves the scaling of the underlying
\acrshort{CC} method, since no coupling between the \acrshort{CC} amplitudes and
multipliers equations is introduced.
The scheme only requires access to a
reference state optimized in the presence of the classical medium and is
thus readily implemented.
The \acrshort{PTE} model offers an efficient approximation to the full
\acrshort{PTED} model with a computational cost comparable to that of
\emph{in vacuo} \acrshort{CC}
theory.\autocite{Olivares_del_Valle1991-of, Begue2005-qn,
Hasegawa2007-jt}

@@

We will now develop a perturbative many-body expansion of the effective
\acrshort{PTED}-\acrshort{CC} Lagrangian \eqref{eq:pted-cc}
The use of a \acrshort{CC} ansatz for the wave function ensures term-by-term
size-extensivity of the energy at all orders.\autocite{Helgaker2000-tz}
The fluctuation potential $\Phi$ is assumed as the perturbation, hence
it will be considered of order 1: $\Phi = O(1)$.
We still assume a closed-shell \acrshort{HF} reference function and
perturbation orders will be counted accordingly.
A superscript index in square brackets - $[i]$ -  will denote an $i$-th
order contribution.
We further assume that real wave functions are used.

@@

\paragraph*{Multipliers}
\begin{equation}%\label{eq:1st-order-mult}
  \bar{\Omega}_{\mu_q}^{[1]} =
    \denom{\mu_q}\tbar{\mu_q}^{[1]}
    + \braket{\mathrm{HF} | \commutator{\Phi}{\cluster{\mu_q}} | \mathrm{HF}}
    + {}^t\p^{[1]}\braket{\mathrm{HF} |
    \commutator{\s}{\cluster{\mu_q}} | \mathrm{HF}}\delta_{\mu_q\mu_1} =0
\end{equation}
and expanding the commutators:
\begin{equation}\label{eq:1st-order-mult}
  \bar{\Omega}_{\mu_q}^{[1]} =
    \denom{\mu_q}\tbar{\mu_q}^{[1]}
    + \braket{\mathrm{HF} | \Phi | \mu_q}
    + {}^t\p^{[1]}\braket{\mathrm{HF} | \s | \mu_q}\delta_{\mu_q\mu_1} =0
\end{equation}
This clearly shows that Eq. \eqref{eq:1st-order-amp}
and Eq. \eqref{eq:1st-order-mult} are complex conjugates. Under the
assumption of real wave functions we can also conclude that:
\begin{equation}
  \tamp{\mu_u}^{[1]} = \tbar{\mu_u}^{[1]},\,\,\forall u
\end{equation}
Thus, as for the amplitudes, despite the closed-shell \acrshort{HF} reference, the singles
multipliers will already appear in first order, due to the
quantum/classical coupling.
Moreover, as is the case \emph{in vacuo}, there are no triples contributions to
the first order amplitudes and multipliers: $\tamp{\mu_3}^{[1]} = \tbar{\mu_3}^{[1]} = 0$.

@@

\paragraph*{Amplitudes}
\begin{equation}\label{eq:2nd-order-amp}
  \begin{aligned}
  \Omega_{\mu_q}^{[2]} &= \denom{\mu_q}\tamp{\mu_q}^{[2]}
  + \braket{\mu_q | \commutator{\Phi}{T^{[1]}} | \mathrm{HF}}
  + {}^t\p^{[2]}\braket{\mu_q | \s | \mathrm{HF}}\delta_{\mu_q\mu_1} \\
  &+ {}^t\p^{[1]}\braket{\mu_q | \commutator{\s}{T^{[1]}} |
  \mathrm{HF}}
  = 0
  \end{aligned}
\end{equation}
As already noted elsewhere,\autocite{Koch1997-nm, Helgaker2000-tz}
the second term in Eq. \eqref{eq:2nd-order-amp} can involve no higher
than triple excitations.
Moreover, the triples first appear to second order and are not
\emph{directly} affected by the quantum/classical coupling, as can be
seen by explicitly considering the singles, doubles and triples
excitation manifolds:
\begin{subequations}
  \begin{align}
  \Omega_{\mu_1}^{[2]} &= \denom{\mu_1}\tamp{\mu_1}^{[2]}
  + \braket{\mu_1 | \commutator{\Phi}{T^{[1]}} | \mathrm{HF}}
  + {}^t\p^{[2]}\braket{\mu_1 | \s | \mathrm{HF}} \nonumber \\
  &+ {}^t\p^{[1]}\braket{\mu_1 | \commutator{\s}{T^{[1]}} |
  \mathrm{HF}}
  = 0 \\
  \Omega_{\mu_2}^{[2]} &= \denom{\mu_2}\tamp{\mu_2}^{[2]}
  + \braket{\mu_2 | \commutator{\Phi}{T^{[1]}} | \mathrm{HF}}
  + {}^t\p^{[1]}\braket{\mu_2 | \commutator{\s}{T_2^{[1]}} |
  \mathrm{HF}}
  = 0 \\
  \Omega_{\mu_3}^{[2]} &= \denom{\mu_3}\tamp{\mu_3}^{[2]}
  + \braket{\mu_3 | \commutator{\Phi}{T_2^{[1]}} | \mathrm{HF}}
  = 0
  \end{align}
\end{subequations}

\paragraph*{Multipliers}
\begin{equation}\label{eq:2nd-order-mult}
  \begin{aligned}
  \bar{\Omega}_{\mu_q}^{[2]} &=
    \denom{\mu_q}\tbar{\mu_q}^{[2]}
    + \braket{\mathrm{HF} | \commutator{\commutator{\Phi}{T^{[1]}}}{\cluster{\mu_q}} | \mathrm{HF}}
    + \sum_{u=1}^\mathcal{M}\braket{\tbar{u}^{[1]} |
    \commutator{\Phi}{\cluster{\mu_q}} | \mathrm{HF}} \\
    &+ {}^t\p^{[2]}\braket{\mathrm{HF} |
    \commutator{\s}{\cluster{\mu_q}} | \mathrm{HF}}\delta_{\mu_q\mu_1}
    + {}^t\p^{[1]}
    \sum_{u=1}^\mathcal{M}\braket{\tbar{u}^{[1]} |
    \commutator{\s}{\cluster{\mu_q}} | \mathrm{HF}}
    =0
  \end{aligned}
\end{equation}
Since first-order singles amplitudes are now nonzero, the double
commutator term will contribute to the singles multipliers equation.
This is markedly in contrast with the derivation \emph{in vacuo}, see
Section \ref{sec:mbpt}.\autocite{Koch1997-nm, Helgaker2000-tz}
For the singles, doubles and triples manifolds the multipliers equations
to second order are given as:
\begin{subequations}
  \begin{align}
    \bar{\Omega}_{\mu_1}^{[2]} &=
      \denom{\mu_1}\tbar{\mu_1}^{[2]}
      + \braket{\mathrm{HF} |
      \commutator{\commutator{\Phi}{T_1^{[1]}}}{\cluster{\mu_1}} | \mathrm{HF}}
      + \sum_{u=1}^\mathcal{M}\braket{\tbar{u}^{[1]} |
      \commutator{\Phi}{\cluster{\mu_1}} | \mathrm{HF}} \nonumber \\
      &+ {}^t\p^{[2]}\braket{\mathrm{HF} |
      \commutator{\s}{\cluster{\mu_1}} | \mathrm{HF}}
      + {}^t\p^{[1]}
      \braket{\tbar{1}^{[1]} |
      \commutator{\s}{\cluster{\mu_1}} | \mathrm{HF}}
      =0 \\
    \bar{\Omega}_{\mu_2}^{[2]} &=
      \denom{\mu_2}\tbar{\mu_2}^{[2]}
      + \sum_{u=1}^\mathcal{M}\braket{\tbar{u}^{[1]} |
      \commutator{\Phi}{\cluster{\mu_2}} | \mathrm{HF}}
      + {}^t\p^{[1]}
      \sum_{u=1}^\mathcal{M}\braket{\tbar{u}^{[1]} |
      \commutator{\s}{\cluster{\mu_2}} | \mathrm{HF}}
      =0 \\
    \bar{\Omega}_{\mu_3}^{[2]} &=
      \denom{\mu_3}\tbar{\mu_3}^{[2]}
      + \sum_{u=1}^\mathcal{M}\braket{\tbar{u}^{[1]} |
      \commutator{\Phi}{\cluster{\mu_3}} | \mathrm{HF}}
      =0
  \end{align}
\end{subequations}
\citeauthor{Koch1997-nm} showed that, assuming real wave functions, the
\emph{in vacuo} second order amplitudes and multipliers equations are
the complex conjugates of each other for the singles, doubles and
triples manifolds.\autocite{Koch1997-nm, Helgaker2000-tz,
Shavitt2009-mr}
This symmetry is a consequence of choosing a closed-shell \acrshort{HF}
determinant as reference function\autocite{Shavitt2009-mr} and is
\emph{broken} by the quantum/classical coupling.
As an example, consider the triples multipliers equation. \emph{In vacuo}
one has:
\begin{equation}\tag{\ref{eq:2nd-order} from Chapter \ref{ch:QM}}
  \denom{\mu_3}\tamp{\mu_3}^{[2]}
  = - \braket{\mu_i | \commutator{\Phi}{T_2^{[1]}} | \mathrm{HF}}
  = - \braket{\tbar{2}^{[1]} | \commutator{\Phi}{\cluster{\mu_3}} |
  \mathrm{HF}}
  = \denom{\mu_3}\tbar{\mu_3}^{[2]},
\end{equation}
whereas now the first-order singles multiplier state also contributes to
the second-order triples multipliers.

@@

From the expansion of the effective Lagrangian we can obtain free energy
corrections up to the desired order. Given the variational nature of the
Lagrangian, only terms fulfilling the $2n+1$ and $2n+2$ rules will
appear in the free energy corrections.
To zeroth-order all variational parameters are zero. To first order we
thus have:
\begin{equation}\label{eq:pted-mbpt-1st}
  G^{[1]} = \braket{\mathrm{HF} | \Phi | \mathrm{HF}}
\end{equation}
The reference energy can thus be recovered as usual:
\begin{equation}
  G_\mathrm{HF} = G^{[0]} + G^{[1]}
\end{equation}
where $G^{[0]}$ includes the quantum/classical coupling.

@@

The second order energy correction will be formed including first order
amplitudes and polarization, while using zeroth order multipliers:
\begin{equation}\label{eq:pted-mbpt-2nd}
  \begin{aligned}
  G^{[2]} &=
  \braket{\mathrm{HF} | \commutator{\Phi}{T^{[1]}} | \mathrm{HF}}
  + \frac{1}{2}{}^t\p^{[1]}\V\p^{[1]} +
  {}^t\p^{[1]}\s_{\mathrm{N}}^{[1]}(\tamp{}^{[1]}, \tbar{}^{[0]}) \\
  &= E^{[2]}
  +
  \frac{1}{2}{}^t\p^{[1]}\V\p^{[1]}
  +
  {}^t\p^{[1]}\braket{\mathrm{HF} | \commutator{\s}{T^{[1]}_1} | \mathrm{HF}}
  = E^{[2]}
  \end{aligned}
\end{equation}
The polarization terms were eliminated by virtue of Eq.
\eqref{eq:1st-order-pol}. Thus, at least formally, $G^{[2]}$ has the same expression as
\emph{in vacuo} and is consistent with \emph{both} electrostatics and
perturbation theory.
We remark that this would not be the case if one had based the
development of \acrshort{MBPT} on the use of a nonlinear
Hamiltonian.\autocite{Angyan1995-co}

The third order energy correction will be formed including first order
amplitudes, multipliers and polarization:
\begin{equation}\label{eq:pted-mbpt-3rd}
  \begin{aligned}
    G^{[3]} &=
    \sum_{u=1}^2\braket{\tbar{u}^{[1]} | \commutator{\Phi}{T^{[1]}} | \mathrm{HF}}
    + {}^t\p^{[1]}\s_{\mathrm{N}}^{[2]}(\tamp{}^{[1]}, \tbar{}^{[1]}) \\
    &=
    E^{[3]}
    + {}^t\p^{[1]}
    \left\lbrace
      \braket{\tbar{1}^{[1]} | \commutator{\s}{T^{[1]}} | \mathrm{HF}}
    + \braket{\tbar{2}^{[1]} | \commutator{\s}{T^{[1]}_2} | \mathrm{HF}}
    \right\rbrace
  \end{aligned}
\end{equation}

>>>>

% mainfile: ../RobertoDiRemigioPhDThesis.tex
%************************************************
\chapter{Electron Correlation and Solvation}\label{ch:solvation-correlation}
\epigraph{Potius sero quam numquam}{
--- \textsc{Titus Livius}, \textit{Ab Urbe Condita}}

@@

The first approximate scheme\footnote{Thanks to prof. Marco Caricato
(University of Kansas) for clarifying some details of his original
derivation of the \gls{PTES} scheme.} truncates the source term to its

@@

   \denom{\mu_q}\tamp{\mu_q} + \braket{\mu_q | \overline{\Phi} | \mathrm{HF}} \nonumber \\
   &+ {}^t\p\braket{\mu_q | \check{\s} | \mathrm{HF}}\delta_{\mu_q\mu_1}

@@

   \denom{\mu_q}\tamp{\mu_q} + \braket{\mu_q | \overline{\Phi} |
   &+ {}^t\p\braket{\mu_q | \check{\s} | \mathrm{HF}}\delta_{\mu_q\mu_1}

@@

\newthought{Both the amplitudes and multipliers equations} are formally identical to

@@

The \acrshort{PTES} approximation can be brought one step further, by
truncating the \acrshort{CCS} expectation value to its multiplier-independent
term already in the effective Lagrangian.
The coupling between the amplitudes and multipliers is broken, while
still improving upon the reference reaction field with correlated
contributions.\autocite{Caricato2011-tx}
The effective Lagrangian would then read as:
\begin{equation}\label{eq:pte-s-cc}
  \begin{aligned}
  \hspace{-20pt}  {}^{\mathrm{PTE(S)}}\lag(\tamp{}, \tbar{},
  \p)_\mathcal{M} &=
  {}^\mathrm{PTE}\lag(\tamp{}, \tbar{})_\mathcal{M} \\
  &+ \frac{1}{2}{}^t\p\V\p + {}^t\p\braket{\mathrm{HF} | \check{\s}_{\mathrm{N}} | \mathrm{HF}}
\end{aligned}
\end{equation}
which uncouples all equations:
\begin{subequations}
  \begin{align}
  {}^{\mathrm{PTE(S)}}\tampEq{\mu_q}(\tamp{}, \tbar{})  &=
   \denom{\mu_q}\tamp{\mu_q} + \braket{\mu_q | \overline{\Phi} | \mathrm{HF}}
   = 0 \\
   {}^{\mathrm{PTE(S)}}\tbarEq{\mu_q}(\tamp{}, \tbar{}, \p) &=
    \denom{\mu_q}\tbar{\mu_q} +
    \braket{\mathrm{HF} | \commutator{\overline{\Phi}}{\cluster{\mu_q}} | \mathrm{HF}}
    \nonumber \\
    &+\sum_{u=1}^{\mathcal{M}}\braket{\tbar{u} |
    \commutator{\overline{\Phi}}{\cluster{\mu_q}} | \mathrm{HF}}
    \nonumber \\
    &+
    {}^t\p\braket{\mathrm{HF} | \commutator{\s}{\cluster{\mu_q}} | \mathrm{HF}}\delta_{\mu_q\mu_1}
    = 0 \\
    {}^{\mathrm{PTE(S)}}\Omega_\p(\tamp{}, \tbar{}, \p)
    &=
    \V\p + \braket{\mathrm{HF} | \check{\s}_{\mathrm{N}} | \mathrm{HF}} = 0
    \label{eq:pte-s-polarization}
  \end{align}
\end{subequations}
The amplitude equations are formally equivalent to those derived in the
\acrshort{PTE} scheme, \emph{vide infra}.
The equations for the multipliers are also formally equivalent to the
\acrshort{PTE} equations, apart from the equation determining the
singles multipliers which features an additional term involving the
polarization:
${}^t\p\braket{\mathrm{HF} | \commutator{\s}{\cluster{\mu_q}} | \mathrm{HF}}\delta_{\mu_q\mu_1}$.
This term is not, however, coupled to the other equations given that
$\p$ is fixed once the singles amplitudes have been determined.

@@

Complete neglect of the polarization functional leads to the most
drastic approximation to the full quantum/classical polarizable
effective Lagrangian, the \acrshort{PTE} scheme.
This is equivalent to differentiating the polarization-independent
\acrshort{PTE} effective Lagrangian:
\begin{subequations}
  \begin{align}
 {}^{\mathrm{PTE}}\tampEq{\mu_q}(\tamp{}, \tbar{}, \p)  &=
   \denom{\mu_q}\tamp{\mu_q} + \braket{\mu_q | \overline{\Phi} | \mathrm{HF}}
    = 0 \\
   {}^{\mathrm{PTE}}\tbarEq{\mu_q}(\tamp{}, \tbar{}, \p)
    &=
    \denom{\mu_q}\tbar{\mu_q} +
    \braket{\mathrm{HF} | \commutator{\overline{\Phi}}{\cluster{\mu_q}}
    | \mathrm{HF}}  \nonumber \\
    &+
    \sum_{u=1}^{\mathcal{M}}\braket{\tbar{u} |
    \commutator{\overline{\Phi}}{\cluster{\mu_q}} | \mathrm{HF}}
    = 0
  \end{align}
\end{subequations}
The \acrshort{PTE} scheme naturally preserves the scaling of the underlying
\acrshort{CC} method, since no coupling between the \acrshort{CC} amplitudes and
multipliers equations is introduced.
The scheme only requires access to a
reference state optimized in the presence of the classical medium and is
thus readily implemented.
The \acrshort{PTE} model offers an efficient approximation to the full
\acrshort{PTED} model with a computational cost comparable to that of
\emph{in vacuo} \acrshort{CC}
theory.\autocite{Olivares_del_Valle1991-of, Begue2005-qn,
Hasegawa2007-jt}

@@

We will now develop a perturbative many-body expansion of the effective
\acrshort{PTED}-\acrshort{CC} Lagrangian.
The use of a \acrshort{CC} ansatz for the wave function ensures term-by-term
size-extensivity of the energy at all orders.\autocite{Helgaker2000-tz}
The fluctuation potential $\Phi$ is assumed as the perturbation, hence
it will be considered of order 1: $\Phi = O(1)$.
We still assume a closed-shell \acrshort{HF} reference function and
perturbation orders will be counted accordingly.
A superscript index in square brackets - $[i]$ -  will denote an $i$-th
order contribution.
We further assume that real wave functions are used.

@@

\paragraph*{Multipliers}
\begin{equation}%\label{eq:1st-order-mult}
  \begin{aligned}
  \bar{\Omega}_{\mu_q}^{[1]} &=
    \denom{\mu_q}\tbar{\mu_q}^{[1]}
    + \braket{\mathrm{HF} | \commutator{\Phi}{\cluster{\mu_q}} | \mathrm{HF}}
    \nonumber \\
    &+ {}^t\p^{[1]}\braket{\mathrm{HF} |
    \commutator{\s}{\cluster{\mu_q}} | \mathrm{HF}}\delta_{\mu_q\mu_1} =0
  \end{aligned}
\end{equation}
and expanding the commutators:
\begin{equation}\label{eq:1st-order-mult}
  \bar{\Omega}_{\mu_q}^{[1]} =
    \denom{\mu_q}\tbar{\mu_q}^{[1]}
    + \braket{\mathrm{HF} | \Phi | \mu_q}
    + {}^t\p^{[1]}\braket{\mathrm{HF} | \s | \mu_q}\delta_{\mu_q\mu_1} =0
\end{equation}
This clearly shows that Eq. \eqref{eq:1st-order-amp}
and Eq. \eqref{eq:1st-order-mult} are complex conjugates. Under the
assumption of real wave functions we can also conclude that:
\begin{equation}
  \tamp{\mu_u}^{[1]} = \tbar{\mu_u}^{[1]},\,\,\forall u
\end{equation}
Thus, as for the amplitudes, despite the closed-shell \acrshort{HF} reference, the singles
multipliers will already appear in first order, due to the
quantum/classical coupling.
Moreover, as is the case \emph{in vacuo}, there are neither triples
amplitudes nor multipliers
to first order: $\tamp{\mu_3}^{[1]} = \tbar{\mu_3}^{[1]} = 0$.

@@

\paragraph*{Amplitudes}
\begin{equation}\label{eq:2nd-order-amp}
  \begin{aligned}
  \Omega_{\mu_q}^{[2]} &= \denom{\mu_q}\tamp{\mu_q}^{[2]}
  + \braket{\mu_q | \commutator{\Phi}{T^{[1]}} | \mathrm{HF}} \\
  &+ {}^t\p^{[2]}\braket{\mu_q | \s | \mathrm{HF}}\delta_{\mu_q\mu_1} \\
  &+ {}^t\p^{[1]}\braket{\mu_q | \commutator{\s}{T^{[1]}} |
  \mathrm{HF}}
  = 0
  \end{aligned}
\end{equation}
As already noted elsewhere,\autocite{Koch1997-nm, Helgaker2000-tz}
the second term in Eq. \eqref{eq:2nd-order-amp} can involve no higher
than triple excitations.
Moreover, the triples first appear to second order and are not
\emph{directly} affected by the quantum/classical coupling, as can be
seen by explicitly considering the singles, doubles and triples
excitation manifolds:
\begin{subequations}
  \begin{align}
  \Omega_{\mu_1}^{[2]} &= \denom{\mu_1}\tamp{\mu_1}^{[2]}
  + \braket{\mu_1 | \commutator{\Phi}{T^{[1]}} | \mathrm{HF}}
  + {}^t\p^{[2]}\braket{\mu_1 | \s | \mathrm{HF}} \nonumber \\
  &+ {}^t\p^{[1]}\braket{\mu_1 | \commutator{\s}{T^{[1]}} |
  \mathrm{HF}}
  = 0 \\
  \Omega_{\mu_2}^{[2]} &= \denom{\mu_2}\tamp{\mu_2}^{[2]}
  + \braket{\mu_2 | \commutator{\Phi}{T^{[1]}} | \mathrm{HF}} \nonumber
  \\
  &+ {}^t\p^{[1]}\braket{\mu_2 | \commutator{\s}{T_2^{[1]}} |
  \mathrm{HF}}
  = 0 \\
  \Omega_{\mu_3}^{[2]} &= \denom{\mu_3}\tamp{\mu_3}^{[2]}
  + \braket{\mu_3 | \commutator{\Phi}{T_2^{[1]}} | \mathrm{HF}}
  = 0
  \end{align}
\end{subequations}

\paragraph*{Multipliers}
\begin{equation}\label{eq:2nd-order-mult}
  \begin{aligned}
  \bar{\Omega}_{\mu_q}^{[2]} &=
    \denom{\mu_q}\tbar{\mu_q}^{[2]}
    + \braket{\mathrm{HF} | \commutator{\commutator{\Phi}{T^{[1]}}}{\cluster{\mu_q}} | \mathrm{HF}}
    + \sum_{u=1}^\mathcal{M}\braket{\tbar{u}^{[1]} |
    \commutator{\Phi}{\cluster{\mu_q}} | \mathrm{HF}} \\
    &+ {}^t\p^{[2]}\braket{\mathrm{HF} |
    \commutator{\s}{\cluster{\mu_q}} | \mathrm{HF}}\delta_{\mu_q\mu_1}
    \nonumber \\
    &+ {}^t\p^{[1]}
    \sum_{u=1}^\mathcal{M}\braket{\tbar{u}^{[1]} |
    \commutator{\s}{\cluster{\mu_q}} | \mathrm{HF}}
    =0
  \end{aligned}
\end{equation}
Since first-order singles amplitudes are now nonzero, the double
commutator term will contribute to the singles multipliers equation.
This is markedly in contrast with the derivation \emph{in vacuo}, see
Section \ref{sec:mbpt}.\autocite{Koch1997-nm, Helgaker2000-tz}
For the singles, doubles and triples manifolds the multipliers equations
to second order are given as:

\begin{subequations}
  \begin{align}
    \bar{\Omega}_{\mu_1}^{[2]} &=
      \denom{\mu_1}\tbar{\mu_1}^{[2]}
      + \braket{\mathrm{HF} |
      \commutator{\commutator{\Phi}{T_1^{[1]}}}{\cluster{\mu_1}} | \mathrm{HF}}
      + \sum_{u=1}^\mathcal{M}\braket{\tbar{u}^{[1]} |
      \commutator{\Phi}{\cluster{\mu_1}} | \mathrm{HF}} \nonumber \\
      &+ {}^t\p^{[2]}\braket{\mathrm{HF} |
      \commutator{\s}{\cluster{\mu_1}} | \mathrm{HF}} \nonumber \\
      &+ {}^t\p^{[1]}
      \braket{\tbar{1}^{[1]} |
      \commutator{\s}{\cluster{\mu_1}} | \mathrm{HF}}
      =0 \\
    \bar{\Omega}_{\mu_2}^{[2]} &=
      \denom{\mu_2}\tbar{\mu_2}^{[2]}
      + \sum_{u=1}^\mathcal{M}\braket{\tbar{u}^{[1]} |
      \commutator{\Phi}{\cluster{\mu_2}} | \mathrm{HF}}
      \nonumber \\
      &+ {}^t\p^{[1]}
      \sum_{u=1}^\mathcal{M}\braket{\tbar{u}^{[1]} |
      \commutator{\s}{\cluster{\mu_2}} | \mathrm{HF}}
      =0 \\
    \bar{\Omega}_{\mu_3}^{[2]} &=
      \denom{\mu_3}\tbar{\mu_3}^{[2]}
      + \sum_{u=1}^\mathcal{M}\braket{\tbar{u}^{[1]} |
      \commutator{\Phi}{\cluster{\mu_3}} | \mathrm{HF}}
      =0
  \end{align}
\end{subequations}
\noindent\citeauthor{Koch1997-nm} showed that, assuming real wave functions, the
\emph{in vacuo} second order amplitudes and multipliers equations are
the complex conjugates of each other for the singles, doubles and
triples manifolds.\autocite{Koch1997-nm, Helgaker2000-tz,
Shavitt2009-mr}
This symmetry is a consequence of choosing a closed-shell \acrshort{HF}
determinant as reference function\autocite{Shavitt2009-mr} and is
\emph{broken} by the quantum/classical coupling.
As an example, consider the triples multipliers equation. \emph{In vacuo}
one has:
\begin{equation}\tag{\ref{eq:2nd-order} from Chapter \ref{ch:QM}}
  \begin{aligned}
  \denom{\mu_3}\tamp{\mu_3}^{[2]}
  &= - \braket{\mu_i | \commutator{\Phi}{T_2^{[1]}} | \mathrm{HF}} \\
  &= - \braket{\tbar{2}^{[1]} | \commutator{\Phi}{\cluster{\mu_3}} |
  \mathrm{HF}}
  = \denom{\mu_3}\tbar{\mu_3}^{[2]},
\end{aligned}
\end{equation}
whereas now the first-order singles multiplier state also contributes to
the second-order triples multipliers.

@@

From the expansion of the effective Lagrangian we can obtain free energy
corrections up to the desired order. Given the variational nature of the
Lagrangian, only terms fulfilling the $2n+1$ and $2n+2$ rules will
appear in the free energy corrections.
To zeroth-order all variational parameters are zero. To first order we
thus have:
\begin{equation}\label{eq:pted-mbpt-1st}
  G^{[1]} = \braket{\mathrm{HF} | \Phi | \mathrm{HF}}.
\end{equation}
The reference energy can thus be recovered as usual:
\begin{equation}
  G_\mathrm{HF} = G^{[0]} + G^{[1]},
\end{equation}
where $G^{[0]}$ includes the quantum/classical coupling.

@@

\arrowthought{The second order energy correction will be formed} including first order
amplitudes and polarization, while using zeroth order multipliers:
\begin{equation}\label{eq:pted-mbpt-2nd}
  \begin{aligned}
  G^{[2]} &=
  \braket{\mathrm{HF} | \commutator{\Phi}{T^{[1]}} | \mathrm{HF}} \\
  &+ \frac{1}{2}{}^t\p^{[1]}\V\p^{[1]} +
  {}^t\p^{[1]}\s_{\mathrm{N}}^{[1]}(\tamp{}^{[1]}, \tbar{}^{[0]}) \\
  &= E^{[2]}
  +
  \frac{1}{2}{}^t\p^{[1]}\V\p^{[1]}
  +
  {}^t\p^{[1]}\braket{\mathrm{HF} | \commutator{\s}{T^{[1]}_1} |
  \mathrm{HF}} \\
  &= E^{[2]}
  \end{aligned}
\end{equation}
The polarization terms were eliminated by virtue of Eq.
\eqref{eq:1st-order-pol}. Thus, at least formally, $G^{[2]}$ has the same expression as
\emph{in vacuo} and is consistent with \emph{both} electrostatics and
perturbation theory.
We remark that this would not be the case if one had based the
development of \acrshort{MBPT} on the use of a nonlinear
Hamiltonian.\autocite{Angyan1995-co}

\arrowthought{The third order energy correction will be formed} including first order
amplitudes, multipliers and polarization:
\begin{equation}\label{eq:pted-mbpt-3rd}
  \begin{aligned}
    G^{[3]} &=
    \sum_{u=1}^2\braket{\tbar{u}^{[1]} | \commutator{\Phi}{T^{[1]}} | \mathrm{HF}}
    + {}^t\p^{[1]}\s_{\mathrm{N}}^{[2]}(\tamp{}^{[1]}, \tbar{}^{[1]}) \\
    &=
    E^{[3]}
    + {}^t\p^{[1]}
      \braket{\tbar{1}^{[1]} | \commutator{\s}{T^{[1]}} | \mathrm{HF}}
      \\
    &+ {}^t\p^{[1]}
     \braket{\tbar{2}^{[1]} | \commutator{\s}{T^{[1]}_2} | \mathrm{HF}}
  \end{aligned}
\end{equation}
